---
title: The Monty Hall Problem
author: SP
date: '2020-10-05'
slug: the-monty-hall-problem
categories:
  - Probability Theory
tags:
  - Paradox
  - Simulation
  - Probability Theory
---


Here is some R code to illustrate the outcomes of the two strategies (staying and switching) discussed in the lecture. You can copy and paste this code to play around with it. 


```{r}
library(ggplot2)
library(dplyr)
library(RColorBrewer)
# number of runs
n = 1000
all.doors = 1:3

# some variables to keep track of waht is going on
win.stay = numeric(n)
win.switch = numeric(n)
chosen = numeric(n)
winning = numeric(n)
openend = numeric(n)

for (i in 1 : n)
{
  # chose a winning door for each time
  winning.door = sample(all.doors,1)
  
  # contest chooses a door 
  chosen.door = sample(all.doors,1)
  
  # game host opens one door, 
  # the remaining door is the one that the contest can switch to
  if(winning.door == chosen.door)
  {
    # we randomly chosen one door that is not the winning door 
    # or the chosen door
    # we open the other one
    # the player can switch to the remaining door
    
    remaining.door = sample(all.doors[-winning.door],1)
    
  }
  
  else 
  {
    # we select the door that is not the winning door
    # and not the chosen door and open it
    opened.door = all.doors[-c(winning.door,chosen.door)]
    
    # the remaining door is then the one that is not chosen
    # and not opened
    remaining.door = all.doors[-c(chosen.door,opened.door)]
    
  }

  win.switch[i] = (remaining.door == winning.door)
  win.stay[i] = (chosen.door == winning.door)
  
  # record for visualization
  chosen[i] = chosen.door
  winning[i] = winning.door

}


print(paste("The strategy 'switch' yielded ", sum(win.switch), " wins."))
print(paste("The strategy 'stay' yielded ", sum(win.stay), " wins."))
```


We can visualize the outcome by plotting when and how often each strategy leads to winning the prize.

```{r, echo=FALSE,out.width = "80%",fig.align="center"}
dat = data.frame(switch = win.switch,stay = win.stay,chosen = chosen,winning = winning)

ggplot(filter(dat,win.stay==T)) + 
  geom_bin2d(aes(x = chosen,y = winning),bins=2.39) + 
  scale_x_continuous(breaks=c(1,2,3),labels=c("Door 1","Door 2","Door 3")) +
  scale_y_continuous(breaks=c(1,2,3),labels=c("Door 1","Door 2","Door 3")) +
  xlab("Initially Chosen") +
  ylab("Prize hiden") +
  ggtitle("Keeping inital door") +
  scale_fill_distiller(name = "Wins",palette=1, direction=1,limits=c(0, 150))  +
  theme_classic() + 
  theme(text = element_text(size = 20))

ggplot(filter(dat,win.switch==T)) + 
  geom_bin2d(aes(x = chosen,y = winning),bins=2.39) + 
  scale_x_continuous(breaks=c(1,2,3),labels=c("Door 1","Door 2","Door 3")) +
  scale_y_continuous(breaks=c(1,2,3),labels=c("Door 1","Door 2","Door 3")) +
  xlab("Initially Chosen") +
  ylab("Prize hiden") +
  ggtitle("Switching door") +
  scale_fill_distiller(name = "Wins",palette=1, direction=1,limits=c(0, 150))  +
  theme_classic() + 
  theme(text = element_text(size = 20))


```
We see two interesting things: (1) "switch" wins exactly when "stay" doesn't win and (2) switching always wins if you chose the wrong door initially. Thus, staying wins with a probability of 1/3 (choosing the right door out of 3), and switching wins with a probability of 2/3. 
